---
title: "OVtool - Omitted Variable tool"
output: github_document
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{courier}
---

**Note: This is a work in progress**

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The <ins>O</ins>mitted <ins>V</ins>ariable tool (`OVtool`) package was designed to assess the sensitivity of research findings to omitted variables when estimating causal effects using propensity score (PS) weighting. This package includes graphics and summary results that will enable a researcher to quantify the impact an omitted variable would have on their results. Burgette et al. (in preparation) describe the methodology behind the primary function in this package, `ov_simgrid()`. This document presents syntax for the implementation of the `ov_simgrid()` function and provides an example of how to interpret the packages' graphical output.

This package is useful in a wide range of applications where researchers want to analyze how sensitive their research findings are to unobserved confounders that were not included in their propensity score and outcome models. It will estimate the potential impact of the unobserved counfounders on both the estimated treatment or exposure effects as well as on the statistical significance of an analysis.

# Example: Synthetic Data 

This package is demonstrated using a random subset of the Global Appraisal of Individual Needs biopsychosocial assessment instrument (GAIN) (Dennis, Titus et al. 2003) from sites that adminstered two different types of substance use disorder treatments (treatment "A" and treatment "B"). The Center for Substance Abuse Treatment (CSAT) funded the sites that administered these two SUD treatments. This dataset consists of 4,000 adolescents. The main goal of this analysis is to understand the effect Treatment A and Treatment B, indicated by `treat`, have on mental health outcomes. 

In this synthetic dataset, there are 2,000 adolescents in each treatment group. Within this dataset there are variables on substance use disorder and mental health outcomes. For this tutorial we are particularly interested in the mental health outcome, `eps7p_3`, emotional problem scale (eps) recorded at three months. Higher values of eps indicate more emotional problems. Substance abuse researchers are particularly interested in whether or not treatment A reduces emotional problems more than treatment B. `eps7p_3` ranges from zero to one, where higher values of EPS indicate more emotional problems. See (Dennis, 2003) for more details on this scale. 

Past research <ins style="color:red">(cite?)</ins> has indicated there are many influential confounders when analyzing adolescents' emotional problems, some included in this synthetic dataset. These variables were measured at baseline: emotional problem scale, adjusted days abstinent (any in past 90) (`ada_0`), substance frequency scale (`sfs8p_0`), substance abuse treatment index (`sati_0`), in recovery (`recov_0`), traumatic stress scale (`tss_0`), and internal mental distress scale (`imds_0`). \newline

We begin by loading the development version of the package from 
[GitHub](https://github.com/) with: 

```{r, begin, echo=T, eval=F}
install_github("jpane24/OVtool")
library(OVtool)
set.seed(24)
```

We can load the synthetic dataset and print to screen the first six observations by running the following two commands:

```{r, data}
data(sud) 
# head(sud)
sud$treat = ifelse(sud$treat == "A", 1, 0)
# get EPS on proper scale according to GAIN
sud$eps7p_0 = sud$eps7p_0/100
sud$eps7p_3 = sud$eps7p_3/100
sud$eps7p_6 = sud$eps7p_6/100
```

The relevant variables in this analysis are:

* **Treatment indicator** `treat`: indicates treatment type where 1 is Treatment "A" and 0 is Treatment "B"

* **Outcome of interest** `eps7p_3`: emotional problem scale at 3-months

* `eps7p_0`: emotional problem scale at baseline

* `sfs8p_0`: substance frequency scale 8-item version at baseline

* `sati_0`: substance abuse treatment index at baseline

* `ada_0`: adjusted days abstinent at baseline

* `recov_0`: indicates whether the adolescent was in recovery at baseline, where 1 is in recovery and 0 is not in recovery

* `tss_0`: traumatic stress scale at baseline

* `mhtrt_0`: mental health treatment in the past 90 days at baseline

* `dss9_0`: depressive symptom scale at baseline

In the next section, we will show how our method works with the average treatment effect (ATE) using a continuous outcome. The OVtool also handles binary outcomes and weights that were estimated using the average treatment effect on the treated (ATT) estimand.

## Continous Outcome: Average Treatment Effect (ATE)

The `OVtool` can either take a vector of weights estimated using any method or a ps object produced by `TWANG`. We begin walking through the OVtool by estimating weights using `ps()` from the `TWANG` package prior to running our outcome model using `OVtool::outcome_model()`. The snippet of code belows walks through an example:

```{r, estTWANG}
## Create Formula
my_formula = as.formula(treat ~ eps7p_0 + sfs8p_0 + sati_0 + ada_0 + recov_0 + 
                          tss_0 + mhtrt_0 + dss9_0)

## Get weights
sud = data.frame(sud)
ps.twang <- twang::ps(my_formula, data = sud, estimand = 'ATE', booster = "gbm",
                      stop.method = "ks.max", verbose=F, ks.exact = T)

# Check Balance
twang::bal.table(ps.twang); # summary(ps.twang)
```

The output produced by the code snippet above shows that TWANG does a reasonable job of balancing. There are additional diagnostics we could check to ensure we have good balance but we move on without diving in further because the purpose of this tutorial is to showcase `OVtool`. The next step is to estimate the treatment effect and analyze the sensitivity of those results using `OVtool`. We first present how a researcher would produce results for their outcomes model (`survey::svyglm()`). There are two options the researcher can take to input the relevant information to get their outcome results using `OVtool::outcome_model()`. 

* Input a `ps.object` from `TWANG` and a `stop.method` (e.g. `"ks.max"`)
or
* Input a vector of `weights`, a data frame containing the `data` used, and the column name representing the treatment indicator, `treatment`. 

The analyst must also provide a column name representing the outcome and a vector of covariates to be included in the final outcome model.

```{r, outcomemodel}
# Get weights (not needed if user inserts a ps object in OVTool)
sud$w_twang = ps.twang$w$ks.max.ATE

# Run Models -- first standardize outcome
sud$eps7p_3_std = sud$eps7p_3/sd(sud$eps7p_3) 

# Use OVtool::outcome_model() to run outcomes model
results = OVtool::outcome_model(ps_object = NULL, # could use ps.twang
                                stop.method = NULL, # could use ks.max
                                data = sud,
                                weights = sud$w_twang, 
                                treatment = "treat",
                                outcome = "eps7p_3",
                                model_covariates = c("eps7p_0", "sfs8p_0", 
                                                     "sati_0", "ada_0",
                                                     "recov_0", "tss_0",
                                                     "mhtrt_0", "dss9_0"),
                                estimand = "ATE")
summary(results$mod_results)
```

The outcome model results show an adjusted treatment effect estimate that accounts for some of the confounding between youth the two treatment programs. From the results, we can see that the effect size is 0.02 (p < 0.427). At this stage, researchers should begin to ask themselves if this effect is real and how sensitive it is. Our tool is used to help answer these sort of logical next step questions. The next snippet of code presents the main function in `OVtool`: `ov_simgrid`. This function requires results from `OVtool::outcome_model` plus additional parameters including:

* `weight_covariates`: a vector of column names representing the covariates used to produce the analysts propensity score weights (these may or may not be the same as the list of covariates used for the outcome model)

* `es_grid`: a vector on an effect size scale representing the association between an unobserved confounder (omitted variable) and the treatment indicator

* `rho_grid`: a vector of correlations to simulate over. These correlations represent the correlation between the omitted variable and the outcome

* `n_reps`: the number of repetitions at each grid point. The package defaults to 101.

The grid, as shown by the x-axis and y-axis in Figure 1 presents the effect size and rho, respectively. We define rho in this setting as the absolute correlation a covariate has with the outcome of interest. Please see Burgette et al. (2020), for additional details.

```{r, ov_simgrid}
# Run OVtool (with weights/not a ps object)
ovtool_results_twang = OVtool::ov_simgrid(model_results=results, 
                                          weight_covariates=c("eps7p_0", "sfs8p_0", 
                                                              "sati_0", "ada_0",
                                                              "recov_0", "tss_0",
                                                              "mhtrt_0", "dss9_0"),
                                          es_grid = NULL,
                                          rho_grid = NULL,
                                          n_reps=2)
# plot(ovtool_results_twang, print_all = "1")
# plot(ovtool_results_twang, print_all = "2")
# plot(ovtool_results_twang, print_all = "3")

```





