---
title: "OVtool - Omitted Variable tool"
output: github_document
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{courier}
---

**Note: This is a work in progress**

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The <ins>O</ins>mitted <ins>V</ins>ariable tool (`OVtool`) package was designed to assess the sensitivity of research findings to omitted variables when estimating causal effects using propensity score (PS) weighting. This package includes graphics and summary results that will enable a researcher to quantify the impact an omitted variable would have on their results. Burgette et al. (in preparation) describe the methodology behind the primary function in this package, `ov_simgrid()`. This document presents syntax for the implementation of the `ov_simgrid()` function and provides an example of how to interpret the packages' graphical output.

This package is useful in a wide range of applications where researchers want to analyze how sensitive their research findings are to unobserved confounders that were not included in their propensity score and outcome models. It will estimate the potential impact of the unobserved counfounders on both the estimated treatment or exposure effects as well as on the statistical significance of an analysis.

# Example: Synthetic Data 

This package is demonstrated using a random subset of the Global Appraisal of Individual Needs biopsychosocial assessment instrument (GAIN) (Dennis, Titus et al. 2003) from sites that adminstered two different types of substance use disorder treatments (treatment "A" and treatment "B"). The Center for Substance Abuse Treatment (CSAT) funded the sites that administered these two SUD treatments. This dataset consists of 4,000 adolescents. The main goal of this analysis is to understand the effect Treatment A and Treatment B, indicated by `treat`, have on mental health outcomes. 

In this synthetic dataset, there are 2,000 adolescents in each treatment group. Within this dataset there are variables on substance use disorder and mental health outcomes. For this tutorial we are particularly interested in the mental health outcome, `eps7p_3`, emotional problem scale (eps) recorded at three months. Higher values of eps indicate more emotional problems. Substance abuse researchers are particularly interested in whether or not treatment A reduces emotional problems more than treatment B. `eps7p_3` ranges from zero to one, where higher values of EPS indicate more emotional problems. See (Dennis, 2003) for more details on this scale. 

Past research <ins style="color:red">(cite?)</ins> has indicated there are many influential confounders when analyzing adolescents' emotional problems, some included in this synthetic dataset. These variables were measured at baseline: emotional problem scale, adjusted days abstinent (any in past 90) (`ada_0`), substance frequency scale (`sfs8p_0`), substance abuse treatment index (`sati_0`), in recovery (`recov_0`), traumatic stress scale (`tss_0`), and internal mental distress scale (`imds_0`). \newline

We begin by loading the development version of the package from 
[GitHub](https://github.com/) with: 

```{r, begin, echo=T, eval=F}
install_github("jpane24/OVtool")
library(OVtool)
set.seed(24)
```

We can load the synthetic dataset and print to screen the first six observations by running the following two commands:

```{r, data}
data(sud) 
# head(sud)
sud$treat = ifelse(sud$treat == "A", 1, 0)
table(sud$treat)
```

The relevant variables in this analysis are:

* **Treatment indicator** `treat`: indicates treatment type where 1 is Treatment "A" and 0 is Treatment "B"

* **Outcome of interest** `eps7p_3`: emotional problem scale at 3-months

* `eps7p_0`: emotional problem scale at baseline

* `sfs8p_0`: substance frequency scale 8-item version at baseline

* `sati_0`: substance abuse treatment index at baseline

* `ada_0`: adjusted days abstinent at baseline

* `recov_0`: indicates whether the adolescent was in recovery at baseline, where 1 is in recovery and 0 is not in recovery

* `tss_0`: traumatic stress scale at baseline

* ``


The package will accept propensity score weights or, if not provided by the user, will calculate propensity score weights using \texttt{TWANG} Ridgeway et al. (2014). In the next section, we will show how our method works with the average treatment effect (ATE) using a continuous outcome. The OVtool also handles binary outcomes and weights that were estimated using the average treatment effect on the treated (ATT) estimand.

